---
title: "Frequentist Inference"
output: 
  html_document:
    keep_md: true
---

```{r setup, include = FALSE}

  knitr::opts_chunk$set(message = FALSE, warning = FALSE)

# Install/load packages

  packs <- c("tidyverse", "knitr", "viridis", "broom", "pwr")
  install.packages(packs[!packs %in% installed.packages()])
  lapply(packs, library, character.only = TRUE)

```

## Part 1. Estimands, estimators, and estimates

The primary aim of a clinical trial is to arrive at some number that tells us something useful about the "true" causal effect of the treatment we are testing. This number, which is calculated from the study data, is called an **estimate** of some corresponding "truth" called an **estimand**. The actual calculation or algorithm we used to provided this estimate of the estimand is called the **estimator**. 

Much of clinical science is about deciding what the important estimands are, asking, "What do we want to know?" On the other hand, much of statistics is about understanding and evaluating the properties of estimators, asking "What is the optimal way to learn it?" 

### A simple example. 

Let's say we want to estimate the mean systolic blood pressure in the population of Irish women, to help plan health services. One way to do this would be to take a random sample from this population, measure their blood pressures, and calculate the mean of the observed values. In this example, the estimand, estimator, and estimate are as follows:

- Estimand - The mean systolic blood pressure in the **population** of Irish women.
- Estimator - The mean systolic blood pressure in our **sample** of this population.
- Estimate - The actual value that arises from our estimator. 

Estimators have different properties that we can use to evaluate how useful they are. Importantly, our understanding of an estimator's properties will usually rely on assumptions, for example, that the sample was truly a random draw from the population. 

To better understand all of this, let's simulate a large number of SBP measurements from a normal distribution with a known mean and standard deviation. This will be our population.  

```{r simulate_population}

# First we will simulate our population

# Set the seed for the random number generator to preserve reproducability
  set.seed(1209) 

  pop_sbp <- rnorm(1e7, 124.5, 18.0) # Simulate a large population of values

  hist(pop_sbp)                      # Plot the distribution of those values
  
  summary(pop_sbp)                   # Summarize the distribution

```

Next, let's draw a random sample from this population, and calculate the mean. The full procedure of taking a random sample of the population and calculating the mean of that sample is our estimator **for** the population mean. The actual number that results is our estimate **of** the population mean. 

```{r sample_1}

  sample_sbp <- sample(pop_sbp, 50, replace = FALSE) # Draw a sample of n = 50

  mean(sample_sbp) # Calculate the mean of the sample

```

We can immediately see that the sample mean is close to, but not exactly, the population mean (which we know was 124.5, from our simulation). Importantly, if we were to repeat the process using the **same** estimator, we would expect a **different** estimate. 

```{r sample_2}

# Draw another sample of n = 50 and calculate the mean
  mean(sample(pop_sbp, 50, replace = FALSE))

```

So what happens if we repeat this process many times? We get a **sampling distribution** of sample means. This concept of a sampling distribution is critical to understanding the **frequentist** school of statistical inference that is most often used to interpret the results of clinical trials, as we will see below.

```{r many_sample_means}

# This is a custom function that drwas a sample from a population and then 
# calcuates the mean of those values. It takes 2 arguments: pop, which is a 
# dataframe object containing the population data we simulated; and n, which is 
# the size of the sample we want to draw.

  rep_means <- function(pop, n, ...){     
    round(mean(sample(pop, n, replace = FALSE)), 2)
  }                                       

# Use this new function 100 times and keep the results from each replicate
  many_sample_means <- replicate(100, rep_means(pop_sbp, 50)) 

  hist(many_sample_means)    # Plot the resulting distribution
  
  summary(many_sample_means) # Summarize it
    

```

### Try it yourself

What do you think would happen to the distribution of the sample mean if we increased the number of observations that we sample from the population? Copy and modify the code above to see what would happen for samples of n = 100, 200, and 500.

## Part 2. Pesky p-values

Now we are going to simulate a clinical trial with a known effect of the intervention, and evaluate our estimate of that effect using the ubiquitous p-value. 

```{r sim_t_test_data}

  set.seed(0236)

# Simulate some study data into a dataframe. Tx effect = 0.5
  study_df <- data_frame(
    y = c(rnorm(20, 0, 1),                       # Control arm outcomes
          rnorm(20, 0, 1) + 0.5),                # Active arm outcomes
    arm = rep(c("Control", "Active"), each = 20) # Arm labels
  )
  
  study_df$arm <- relevel(factor(study_df$arm), ref = "Control")

# Plot the distribtion of outcomes in each arm
  ggplot(study_df, aes(x = y, fill = arm)) +
    geom_histogram() +
    facet_wrap(~arm, ncol = 1) +
    scale_fill_viridis(guide = FALSE, discrete = TRUE, end = 0.85)
  
# Note: When we plotted the distribution before, we used so-called "base" R 
# plotting. Going forward, we will more often used a package called ggplot2 
# (which is included in the tidyverse family of packages).  

```

Now we will pretend that we don't know what the true effect of the treatment is (0.5). The difference in means between the two arms in this sample is `r round(mean(study_df$y[study_df$arm == "Active"]) - mean(study_df$y[study_df$arm == "Control"]), 2)`. This of course isn't exactly 0.5, since the tx effect is added to a variable outcome that is measured in a random sample. Thus some sampling variability should be expected.  

The actual test statistic we will use is called Student's t. It is a ratio of this difference in means to the standard error (which is what we, somewhat confusingly, call the standard deviation of a sampling distribution). You can think of it as a "signal to noise ratio".  

Let's do it "by hand" so we can see how the p-value is calculated. 

The first step is to calculate the value for t from our sample data. This is our estimate. 

```{r calculate_t}

# https://en.wikipedia.org/wiki/Welch's_t-test

# First, just to see how eveCalcuate T "by hand"
  m_act <- mean(study_df$y[study_df$arm == "Active"])
  m_con <- mean(study_df$y[study_df$arm == "Control"])
  v_act <-  var(study_df$y[study_df$arm == "Active"])
  v_con <-  var(study_df$y[study_df$arm == "Control"])
  n <- 20 # Sample size in each group
  se <- sqrt(v_act/n + v_con/n)
  df_s <- (v_act/n + v_con/n)^2 / # Need this for the sampling dist. below
          ((v_act^2 / (n^2 * (n - 1))) + 
           (v_con^2 / (n^2 * (n - 1))))
  
  t_sample <- round((m_act - m_con) / se, 2)
  
  t_sample

```

Then we want to plot the sampling distribution of t under the null hypothesis that there is no different in the means (i.e. t = 0). The key parameter for this distribution is the degrees of freedom we calculated above (df_s). 

```{r t_null_dist}

# Get the expected sampling distibution under a null hypotheis of no difference
  
  g1 <- ggplot(data_frame(x = c(-4 , 4)), aes(x = x)) + 
    stat_function(fun = dt, args = list(df = df_s), 
                  geom = "area", fill = viridis(1), alpha = 0.3) +
    xlim(c(-4 , 4)) +
    xlab("t") +
    ylab("Density") +
    theme_minimal()

  g1

```

Then we plot the position of t for our sample, and calculate the area of the t distribution (under the null) for values of t as big or bigger than the value we actually observed. 

```{r t_one_sided}

  g2 <- g1 +
    geom_vline(xintercept = t_sample) +
    stat_function(xlim = c(t_sample , 4), 
                  fun = dt, args = list(df = df_s), 
                  geom = "area", fill = viridis(1)) +
    scale_x_continuous(breaks = c(-3, -1.5, 0, 1.5, t_sample, 3)) +
    theme(panel.grid.minor = element_blank()) +
    ggtitle(paste0("The proportion of the total area in the darker part of the distribution\n for t (given the null is true) is ", signif(pt(t_sample, df_s, lower.tail = FALSE), 2)))

  g2

```

This means that **IF** there was no treatment effect, **AND** the assumptions underlying our calculation of t distribution under the null hold, **AND** we were able repeat our experiment many, many times, **THEN** we would only expect to see a value as large or larger than the one we calculated in *our* sample, 1.2% of the time. This is our p-value. Please note that there is no way to correctly explain what this p-value is in a single sentence. 

So what can we do with this p-value? That's where things start to get tricky. At the risk of over-simplifying things, Fisher saw the p-value as a continuous measure of how surprised we should be if the data were truly generated under a mechanism of "no effect". So then a small p-value indicates there might be an effect worth further exploring or looking for in continued experiments. Then, if you were able able to repeat an experiment several times and consistently produce a small p-value, you might finally conclude that your intervention "works". 

Others would use the p-value as a hypothesis test. In the example above, it would be the null hypothesis of t ≤ 0 vs the alternative of t > 0. Then you would set some threshold for p (e.g. 5%) for which you would "reject the null". We'll come back to this way of thinking when we discuss power, and type 1 and 2 errors below.    

In the example above, we only considered our estimate of t with respect to one tail of the sampling distribution of t under the null. This is what we refer to as a one-sided test. However, it usually makes more sense to make it relative to both tails, so that the null is t = 0 vs. t < 0 or t > 0. 

```{r t_two_sidedd}

  g3 <- g2 +
    stat_function(xlim = c(t_sample , 4), 
                  fun = dt, args = list(df = df_s), 
                  geom = "area", fill = viridis(1)) +
    stat_function(xlim = c(-4, -t_sample), 
                  fun = dt, args = list(df = df_s), 
                  geom = "area", fill = viridis(1)) +
    ggtitle(paste0("The proportion of the total area in the darker part of the distribution\n for t is ", signif(2 * pt(t_sample, df_s, lower.tail = FALSE), 3)))

  g3

```

Now this was a fair bit of work just to do a simple t-test. So now let's R do this for us (and confirm our result).

```{r t_test}

  t_1 <- t.test(y ~ arm, data = study_df)

  t_1

```

### Side-note: Linear models

We noted in EH6124 that most commonly used statistical tests can be represented by linear models. The advantage of this is two-fold. First, linear models can incorporate covariate information which often leads to more efficient estimators by integrating out variability in the outcome predicted by the covariates. It's also useful to think in terms of linear models because it's much easier than trying to remember a multitude statistical tests. 

For example, the t-test is used when we are interested in a difference in means and when we have a fairly small sample size (the specific problem the t-test addresses is that when we have a small sample and we need to use the sample standard deviation to work our the distribution of the sample mean, then we should have more uncertainty in that distribution, i.e. it should be wider - it should have "fatter tails").

```{r linear_model}

  lm_1 <- lm(y ~ arm, data = study_df)

  pvalue <- filter(tidy(lm_1), term == "armActive")$p.value
  
  summary(lm_1)

```

So here we get a p-value of `r signif(pvalue, 3)` vs. `r signif(t_1$p.value, 2)` from the t-test. Of course the difference in means, the effect estimate, is exactly the same (`r round(t_1$estimate[1] - t_1$estimate[2], 2)`). It's the sampling distributions of the two estimators of that effect (t vs the regression coefficient) that are different.

## Part 3. Error control and power

In the example above, we focused mostly on Fisher's interpretations of the  p-value given by a particular set of set of observations as a continuous measure of evidence for/against a null hypothesis/model (though he still viewed being able to repeatedly procure a small p-value as an important step before declaring the existence of an experimental effect). However, this idea was met with resistance from the very beginning. Much of this resistance came from the Neyman-Pearson (NP) school of thought, whose proponents instead thought that p-values should be used to conduct the binary hypothesis tests we noted above. 

When we choose to use such tests, we are using a relaxed version of the following logic (called [modus tollens](https://en.wikipedia.org/wiki/Modus_tollens)):

If A, then B | Not B, then not A

These hypothesis tests are used to make a similar argument: **If** the null hypothesis/model is true, **then** the data should be more likely to look one way. If however the data **don't** look that way, **then** we should consider rejecting the null. 

This means we can incorrectly reject the null then it's true; or incorrectly fail to reject the null when it isn't. From this point of view, our goal when using statistics is to control the probability of making these errors. 

Let's look at some simulations. 

First we will generate some data under a null, normal model, with a random allocation to study arms. 

```{r null_data}

  n <- 100

  data <- data_frame(
    sbp = rnorm(n, 124.5, 18.0), 
    arm = sample(c("Active", "Control"), size = n, replace = TRUE)
  )
  
# Plot the distribtion of outcomes in each arm
  ggplot(data, aes(x = sbp, fill = arm)) +
    geom_histogram() +
    facet_wrap(~arm, ncol = 1) +
    scale_fill_viridis(guide = FALSE, discrete = TRUE, end = 0.85)
  
# Regress SBP on the treatment arm  
  summary(lm(sbp ~ arm, data = data))

```

Now let's look at what happens if we repeat this process many times.

```{r null_data_repeat}

  make_pvalues <- function(n, ...){

    data <- data_frame(
      sbp = rnorm(n, 124.5, 18.0), 
      arm = sample(c("Active", "Control"), size = n, replace = TRUE)
    )
    
    data$arm <- relevel(factor(data$arm), ref = "Control")

    result <- tidy(lm(sbp ~ arm, data = data)) %>%
      filter(term == "armActive")
    
    return(result$p.value)

  }

  k <- 2000
  results <- as_data_frame(replicate(k, make_pvalues(100))) %>%
    rename(p_value = value)

  ggplot(results, aes(x = p_value)) +
    geom_histogram() +
    geom_vline(xintercept = 0.05, color = "red", linepype = "dashed")
  
```

As some of you might have expected, about 5% of the p-values fall below the nominal 5% level (`r round(table(cut(results$p_value, c(0, 0.05, 1)))[1] / k, 3) *100`% in this set of simulations)

Just for fun, what will happen to the distribution of p-values if I decrease the sample size of the individual replicates?

```{r p_distribution_null_1}

  k <- 2000
  results <- as_data_frame(replicate(k, make_pvalues(20))) %>% # n = 20
    rename(p_value = value)

  ggplot(results, aes(x = p_value)) +
    geom_histogram() +
    geom_vline(xintercept = 0.05, color = "red", linepype = "dashed")
  
```

Again, about 5% of the p-values fall below the nominal 5% level (`r   round(table(cut(results$p_value, c(0, 0.05, 1)))[1] / k, 3) *100`% - the difference is just simulation error - if we ran more simulations, this would be 5% for sure)

And if I increase the sample size?

```{r p_distribution_null_2}

  k <- 2000
  results <- as_data_frame(replicate(k, make_pvalues(1000))) %>% # n = 1000
    rename(p_value = value)

  ggplot(results, aes(x = p_value)) +
    geom_histogram() +
    geom_vline(xintercept = 0.05, color = "red", linepype = "dashed")
  
```

The same result (`r round(table(cut(results$p_value, c(0, 0.05, 1)))[1] / k, 3) * 100`%). To put a finer point on this, **when the data are indeed generated by a null model**, the distribution of p-values is constant; and because p is based on a sampling distribution, which is in turn partly determined by the sample size (via the standard error), then 5% of p values will be below the 5% threshold, regardless of sample size. 

The other important addition from the NP view is the introduction of some alternative hypothesis, so that when we reject the null (based on a small p-value) then we are implicitly accepting some alternative. Thus we can restate the two types of possible error we noted above: 

- We can reject the null when it's true, thus declaring there is an effect (the alternative) when there isn't: a false positive. 

- We can accept the null when it's false, thus declaring there is no effect (the alternative) when there actually is: a false negative. 

The first of these is what we just looked at above, when we simulated data under the null and say how many times we would reject it when it was in fact true - and that we could "control" this error rate using the p-value. If I'm happy accepting making this kind of mistake 5% of the time, then my p-value for "significance" should be 0.05. If I want to make that more or less stringent, I can. However, this only really makes sense if I am also happy thinking of what I'm doing as one experiment in a long succession of such experiments, like a machine making widgets. 

So now let's look at the second kind of error, and generate some data with a known effect. 

```{r alt_data_repeat}

  make_pvalues <- function(n, effect, ...){

    data <- data_frame(
      sbp = rnorm(n, 124.5, 18.0), 
      arm = sample(c("Active", "Control"), size = n, replace = TRUE)
    )
    
    data$arm <- relevel(factor(data$arm), ref = "Control")

    
    data$sbp[data$arm == "Active"] <- data$sbp[data$arm == "Active"] + effect
  
    result <- tidy(lm(sbp ~ arm, data = data)) %>%
      filter(term == "armActive")
    
    return(result$p.value)

  }

  k <- 2000
  results <- as_data_frame(replicate(k, make_pvalues(100, -2))) %>%
    rename(p_value = value)

  ggplot(results, aes(x = p_value)) +
    geom_histogram() +
    geom_vline(xintercept = 0.05, color = "red", linepype = "dashed")
  
```

Now if I use p = 0.05 to reject the null, which is now the correct decision (since I know I simulated data with "an effect"), I will do so `r round(table(cut(results$p_value, c(0, 0.05, 1)))[1] / k * 100, 3)`% of the time. That is my **power**. Alternately, I will accept the null (wrongly) `r 100 - round(table(cut(results$p_value, c(0, 0.05, 1)))[1] / k * 100, 3)`% of the time. That is my type 2 error rate. 

Most of the time people recommend a power of at least 80% (more on this later), so clearly this is a very low power. That's because our effect (= -2) is small relative to the variability in the outcome. So we need a bigger effect or a bigger sample to overcome that variability and increase the power. So let's greatly increase the sample size. 

```{r alt_data_2}

  k <- 2000
  results <- as_data_frame(replicate(k, make_pvalues(2000, -2))) %>%
    rename(p_value = value)

  ggplot(results, aes(x = p_value)) +
    geom_histogram() +
    geom_vline(xintercept = 0.05, color = "red", linepype = "dashed")

```

Whoa! What just happened? Now if I use p = 0.05 to (correctly) reject the null, my power is `r round(table(cut(results$p_value, c(0, 0.05, 1)))[1] / k * 100, 3)`%, and my type 2 error rate is `r 100 - round(table(cut(results$p_value, c(0, 0.05, 1)))[1] / k * 100, 3)`%. 
 
Here is an important point - there is no such thing as **A** study's power. Power is specific to both a study's design, and **AN** effect size. Thus the same study will produce a different power for different effect sizes. In these simulated examples, we **know** what the alternative hypothesis is (effect = -2), while for a real study, there are many plausible alternative hypotheses (i.e. many plausible potential effect sizes) and we can't know which is true, or else we wouldn't need to run the study. 

However, one of these hypothetical effect sizes stands out among all the rest. It's the one Senn refers to as the smallest possible effect that we'd like to not miss. In other words, many of these plausible effect sizes are so small that we don't really care if they exist or not (when we are talking about clinical research where the goal is to intervene on people), but there are other, larger effect sizes we certainly would like to be able to detect with a high power - and if we design the study to have high power for the smallest of these, then we will have high power to detect all of them. So let's plan a study to detect an effect = 2, given what we know about the distribution of the outcome we used in the simulation. 

```{r power_calc}
   
   effect_size <- -2
   sd <- 18
   d <- effect_size / sd

   pwr.t.test(d = d, sig.level = .05, power = 0.90)
# I am being a bit lazy here, using the power based on a t-test when I know
# that I am going to analyze the data with a linear model - but sample sizes are
# an educated guess at best - so close enough for now!   

```

Ha! So to detect an effect this small (relative to the SD of the SBP distribution), I need 3400 people in my study for achieve a power of 90% (i.e. to produce a study that will detect this effect, based on a binary hypothesis test vs the null, 90% of the time). Now, let's confirm this with a simulation. 

```{r power_sim}

  k <- 2000
  results <- as_data_frame(replicate(k, make_pvalues(3400, 2))) %>%
    rename(p_value = value)

  ggplot(results, aes(x = p_value)) +
    geom_histogram() +
    geom_vline(xintercept = 0.05, color = "red", linepype = "dashed")

```

From the above I can see that `r round(table(cut(results$p_value, c(0, 0.05, 1)))[1] / k * 100, 3)`% of p-values from the simulations fall below 5% - that's my power, which is just what I expected (with a bit of simulation error). 

I want to point out one more thing, which is the distribution of p-values that fall under 0.05 when the alternative is in fact true and we have a high power test, as we just simulated.

```{r alt_data_le_05}

  filter(results, p_value <= 0.05) %>%
  ggplot(aes(x = p_value)) +
    geom_histogram() +
    geom_vline(xintercept = 0.05, color = "red", linepype = "dashed")

```

So how many of those p-values are close to 0.05? Very few! What does this mean? It means that when some alternative effect is true and the trial is very well powered to detect that effect, the distribution of p-values is heavily skewed towards very small values, values much smaller than 0.05. This is why many people are skeptical of papers and bodies of literature that report many "just significant" p-values, as it suggests publication bias and p-hacking. 

